---
title: "gstream_width"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Load the data and select the first data (north and south walls) as an example.

```{r}
suppressPackageStartupMessages({
  library(sf)
  library(dplyr)
  library(gstream)
  library(rnaturalearth)
  library(units)
  library(ggplot2)
})
coast = rnaturalearth::ne_coastline(scale = "medium", returnclass = "sf")

All = gstream::read_usn()
BB = sf::st_bbox(All)
x = All |>
  dplyr::slice_max(date)
```


Now make a somewhat arbitrary box.
```{r}
west = -71
east = -61
north = 45
south = 32
bb = sf::st_bbox(c(xmin = west, ymin = south, xmax = east, ymax = north),
                 crs = sf::st_crs(All)) |>
  sf::st_as_sfc()



plot(x['wall'], pch = ".", axes = TRUE, reset = FALSE, extent = BB, key.pos = NULL)
plot(bb, add = TRUE, lty = "dashed")
plot(sf::st_geometry(coast), add = TRUE)
```



Now we crop the wall data to the box.  Plotted below shows that the lines are both oriented eastward (first point of each plotted with circle symbol.)

```{r, warning = FALSE}
z = crop_usn(x, bb) 
p = st_cast(z, "POINT")

plot(z['wall'], reset = FALSE, extent = BB)
plot(p['wall'] |> dplyr::filter(wall == 'north') |> dplyr::slice(1), add = TRUE, col = "black")
plot(p['wall'] |> dplyr::filter(wall == 'south') |> dplyr::slice(1), add = TRUE, col = "black")
plot(bb, add = TRUE, lty = "dashed")
plot(sf::st_geometry(coast), add = TRUE)
```


Now merge the two lines into a closed polygon.

```{r, warning = FALSE}
poly = walls_to_polygons(z)


plot(z['wall'], reset = FALSE, extent = BB)
plot(poly[1], add = TRUE)
plot(z['wall'], add = TRUE)
plot(p['wall'] |> dplyr::filter(wall == 'north') |> dplyr::slice(1), add = TRUE, col = "black")
plot(p['wall'] |> dplyr::filter(wall == 'south') |> dplyr::slice(1), add = TRUE, col = "black")
plot(bb, add = TRUE, lty = "dashed")
plot(sf::st_geometry(coast), add = TRUE)
```


Let's take a bigger slice of the dataset.  Below we can see how the process can be streamlined (no pun intended).  We'll add a `month` category for subsequent plotting.

```{r, warning = FALSE}
VERBOSE = TRUE
p = read_usn() |>
  crop_usn(bb) |>
  walls_to_polygons() |>
  dplyr::mutate(month = factor(format(date, "%b"), levels = month.abb),  .after = 1)
p
```

Let's plot up some time series.

```{r, warning = FALSE}
ggplot(data = p, aes(x = date, y = area)) +
  geom_point(alpha = 0.25) + 
  labs(y = "Boxed Area", title = "Gulf Stream Boxed Area") + 
  facet_wrap(~month)
```


```{r, warning = FALSE}
ggplot(data = filter(p, yc < 100), aes(x = date, y = yc)) +
  geom_point(alpha = 0.25) + 
  labs(y = "Boxed Centroid Latitude", title = "Gulf Stream Boxed Centroid") + 
  facet_wrap(~month)
```



