---
title: "gstream"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Serving Gulf Stream datasets directly from R.

## Requirements

[R v4.1+](https://www.r-project.org/)

[rlang](https://CRAN.R-project.org/package=rlang)

[dplyr](https://CRAN.R-project.org/package=dplyr)

[sf](https://CRAN.R-project.org/package=sf)


## Installation

Use the [remotes](https://CRAN.R-project.org/package=remotes) package to install directly from github.

```         
remotes::install("BigelowLab/gstream)
```

## Data from US Navy

### Arhcived data

[NOAA's Ocean Prediction Center](https://ocean.weather.gov/) provides a FTP server](https://ftp.opc.ncep.noaa.gov/grids/experimental/GStream) for downloads by year.  We have downloaded these and repackaged into spatial format files - these are included with the `gstream` package. They also provide [daily updates](https://ocean.weather.gov/gulf_stream_latest.txt).

```{r, warn = FALSE}
suppressPackageStartupMessages({
  library(sf)
  library(dplyr)
  library(gstream)
  library(rnaturalearth)
})

x = read_usn() |>
  dplyr::glimpse()
```

This reads in all of the data stored with the package. We can then do a simple plot of all of the locations.

```{r}
bb = sf::st_bbox(x)
coast = rnaturalearth::ne_coastline(scale = "medium", returnclass = "sf")

plot(x['wall'], pch = 3, axes = TRUE, reset = FALSE)
plot(sf::st_geometry(coast), add = TRUE)
```

### Downloading daily updates and configuration

**Note** that you don't need to create the configuration file if you are not downloading data.

The daily data is hosted by by [NOAA's Ocean Prediction Center](https://ocean.weather.gov/) In particular they post the US Navy's [daily Gulf Stream point data](https://ocean.weather.gov/gulf_stream_latest.txt) for the north and south walls. These can be downloaded.  We provide a mechanism for storing the URL of the daily data, the path to where you want to store the downloads and a simple script for downloading.  The configuration can be stored anywhere, but by default we look for it isn `~/.gstream`.


```{r}
cfg = read_configuration()
cfg
```

Obviously, you will want to modify the `rawpath` to suit your own needs. We then set up a cron job to make the daily download at local 6pm.

```
# gstream data
0 18 * * * /usr/local/bin/Rscript /Users/ben/Library/CloudStorage/Dropbox/code/projects/gsi/inst/scripts/usn_daily_download.R >> /dev/null 2>&1
```


### Ordering USN data

The USN data is not ordered, that is the points for a given day are not following a polyline. 

```{r}
d = dplyr::filter(x, date == as.Date("2020-12-19"), wall == "north")
plot(sf::st_geometry(d), type = "l", axes = TRUE)
```


But we **attempt** to order them.  We have a start, but clearly there are cases where something goes awry.

```{r}
do = order_usn(d)
plot(sf::st_geometry(do), type = "l", axes = TRUE)
```

Sometimes it works.

```{r}
d = dplyr::filter(x, date == as.Date("2020-01-03"), wall == "north")
do = order_usn(d)
plot(sf::st_geometry(d), type = "l", axes = TRUE, reset= FALSE)
plot(sf::st_geometry(do), type = "l", add = TRUE, col = "orange")
```

